---
title: "Receptor Assay Data Tidying"
author: "Alicia Rich"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

library(tidyverse)
library(forcats)
library(conflicted)
library(plotly)
library(htmltools)
library(htmlwidgets)
library(drc)
library(shiny)
library(bslib)
library(flexdashboard)
library(paletteer)
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = ""
  )

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(plotly::layout)

assays <- read_csv("data/receptors/esr_assays_revised.csv")

```

# Prepare Data

## Load Assay Data

- Ensure dose values are numeric and in consistent units. If needed, convert doses to a common unit (e.g., molar concentrations for chemicals, or mg/mL for extracts) or add a column for the log10 of dose for convenience.
- Check that each receptor–treatment combination has the expected number of replicates and dose points. If any treatment has only a single dose (or only the control), mark it for special handling (since a dose–response curve cannot be fit with a single point).

```{r}
treatments <- c(
   control               = "Control"        ,
   vehicle_MeOH          = "Veh MeOH"       ,
   vehicle_DMSO          = "Veh DMSO"       ,
   standard_estradiol    =  "E2"            ,
   standard_estrone      =  "E1"            ,
   standard_estriol      =  "E3"            ,
   psm_genistein         =  "GEN"           ,
   psm_daidzein          =  "DAID"          ,
   psm_coumestrol        =  "COU"           ,   
   microbial_equol       =  "EQUOL"         ,
   microbial_enterodiol  =  "ENTDIOL"       ,
   diet_biscuit_5M1S     =  "LS Cin"        ,
   diet_biscuit_5MA4     =  "Browse Biscuit",
   diet_biscuit_5M1G     =  "LS Ban"        ,
   diet_biscuit_5MA3     =  "High Fiber"    ,
   diet_biscuit_5M6C     =  "Insect"        ,
   diet_biscuit_5M02     =  "Leafeater"     ,
   diet_biscuit_5M02     =  "LE Primate"    ,
   diet_biscuit_5MA2     =  "Pellet OSU"    ,
   diet_lettuce          =  "Lettuce"       ,
   unknown_random        =  "Random"        ,
   unknown_wi            =  "WI"        
)

receptor_labs <- tibble(
  labs_er    = c(rep("ER\u03B1", 3), rep("ER\u03B2", 3)),
  labs_genus = rep(c("<i>Gorilla</i>", "<i>Xanthonycticebus</i>", "<i>Homo</i>"), 2),
  receptor   = c("gorilla_alpha", "loris_alpha", "human_alpha", "gorilla_beta", "loris_beta", "human_beta")
) 

treatment_labs <- tibble(
  labs_treatment = c(
    "E2"            ,
    "E1"            ,
    "E3"            ,
    "Genistein"     ,
    "Daidzein"      ,
    "Coumestrol"    ,
    "Equol"         ,
    "Enterodiol"    ,
    "Biscuit (5M1S)",
    "Biscuit (5MA4)",
    "Biscuit (5M1G)",
    "Biscuit (5MA3)",
    "Biscuit (5M6C)",
    "Biscuit (5M02)",
    "Biscuit (5M02)",
    "Biscuit (5MA2)",
    "Lettuce"       ,
    "Random?"       ,
    "WI" 
  ),
  groups_treatment = c(
    rep("Estrogen", 3),
    rep("PSM", 3),
    rep("Microbial", 2),
    rep("Diet Extract", 11)
  ),
  treatment      = c(
    "standard_estradiol"  ,
    "standard_estrone"    ,
    "standard_estriol"    ,
    "psm_genistein"       ,
    "psm_daidzein"        ,
    "psm_coumestrol"      ,
    "microbial_equol"     ,
    "microbial_enterodiol",
    "diet_biscuit_5M1S"   ,
    "diet_biscuit_5MA4"   ,
    "diet_biscuit_5M1G"   ,
    "diet_biscuit_5MA3"   ,
    "diet_biscuit_5M6C"   ,
    "diet_biscuit_5M02"   ,
    "diet_biscuit_5M02"   ,
    "diet_biscuit_5MA2"   ,
    "diet_lettuce"        ,
    "unknown_random"      ,
    "unknown_wi"          
  )             
) %>% mutate(treatment = factor(treatment))

```

```{r}
esr_data <- assays %>%
  mutate(Treatment = if_else(Dose == "Control" |  str_detect(Dose, "Veh"), Dose, Treatment)) %>%
  mutate(receptor  = factor(as.character(str_glue("{Species}_{ER}"))),
         treatment = fct_recode(Treatment, !!!treatments), .keep = "unused") %>%
  arrange(Date, receptor) %>%
  mutate(plate     = factor(consecutive_id(Date, receptor))) %>%
  mutate(dose_type = case_when(
           str_detect(Dose, "mg/ml") ~ "mg/ml",
           str_detect(Dose, "Veh")   ~ "vehicle",
           Dose == "Control"         ~ "control",
           .default = "molar"
           ),
         dose = as.numeric(str_remove_all(Dose, "[^\\d|-|\\.]"))
         ) %>%
  mutate(dose = if_else(dose_type == "molar", as.numeric(str_glue("1e{Dose}")), dose),
         dose_type = factor(dose_type)) %>%
  arrange(plate, receptor, treatment, dose) %>%
  mutate(Dose = factor(Dose)) %>%
  mutate(dil_series = consecutive_id(Dose), .by = c(plate, treatment)) %>%
  select(receptor,
         plate,
         Date,
         treatment,
         dose_type,
         dil_series,
         dose,
         starts_with("luciferase"),
         starts_with("bgal")
         ) %>%
  pivot_longer(
    c(starts_with("luciferase"), starts_with("bgal")),
    names_pattern = c("(^\\w+)_(\\d$)"),
    names_to     = c("value_type", "replicate"),
    values_to    = "value"
    ) %>%
  mutate(dil_series = if_else(is.na(dose), row_number(), dil_series), .by = c(receptor, plate, treatment, value_type))  %>%
  pivot_wider(
    names_from  = value_type,
    values_from = value
  ) %>%
  rowwise() %>%
  mutate(resp_norm = luciferase/bgal) %>%
  ungroup() %>%
  mutate(veh_norm = mean(resp_norm[str_starts(treatment, "vehicle")]), 
         con_norm = mean(resp_norm[treatment == "control"]),
         .by = "plate") %>%
  filter(!is.na(dose)) %>%
  rowwise() %>%
  mutate(fold_act = resp_norm/veh_norm) %>%
  ungroup() %>%
  mutate(log_dose = log10(dose)) %>%
  left_join(receptor_labs, by = "receptor") %>%
  left_join(treatment_labs, by = "treatment") %>%
  mutate(group_plate = factor(as.character(str_glue("{as.character(receptor)}_{as.character(plate)}_{as.character(treatment)}"))),
         group_pooled = factor(as.character(str_glue("{as.character(receptor)}_{as.character(treatment)}")))) %>%
  select(
    labs_genus,
    labs_er,
    receptor,
    plate,
    group_plate,
    group_pooled,
    groups_treatment,
    labs_treatment,
    treatment,
    dose_type,
    dil_series,
    log_dose,
    dose,
    replicate,
    luciferase,
    bgal,
    resp_norm,
    fold_act
  )

write_csv(esr_data, "data/receptors/esr_assays_tidy.csv")
```

## Select the best fit Model

```{r}
names_2 <-  c("slope","ED50")
names_3 <-  c("slope","lower","upper")
names_4 <-  c("slope","lower","upper","ED50")
names_5 <-  c("slope","lower","upper","ED50", "Asymmetry")

candidates <- list(
  LL.2(names = names_2),    # 2-parameter log-logistic (fixed lower=0, fixed upper=1)
  LL.3(names = names_3),    # 3-parameter log-logistic (estimate upper asymptote)
  LL.4(names = names_4),    # 4-parameter log-logistic (estimate both asymptotes + slope + ED50)
  LL.5(names = names_5),    # 5-parameter log-logistic (adds curve asymmetry)
  W1.4(names = names_4),    # 4-parameter Weibull
  W2.4(names = names_4)     # alternative 4-parameter Weibull
)
```


```{r}
data_plate <- esr_fit %>%
  select(
    group = group_plate,
    log_dose,
    dose,
    fold_act
  ) %>%
  mutate(n_doses = n_distinct(dose), .by = "group") %>%
  filter(n_doses > 2) %>%
  select(-n_doses) %>%
  group_by(group) %>%
  group_split() %>%
  set_names(., map_depth(., 1, \(x) as.character(first(x$group))))

drc_plates <- imap(data_plate, function(d, nm) {
  assign(nm, d, envir = .GlobalEnv)        
  drm(fold_act ~ dose,
      data = get(nm, envir = .GlobalEnv), 
      fct  = LL.4(names = names_4))
})
```

```{r warning=FALSE, message = FALSE, comment = "", echo = FALSE}
msel_plates <- list()
for (nm in names(drc_plates)) {
  msel_plates[[nm]] <- mselect(drc_plates[[nm]], fctList = candidates)
}
```


```{r}
e2_references <- esr_data %>%
  filter(treatment == "standard_estradiol") %>%
  select(
    group = plate,
    log_dose,
    dose,
    fold_act
  ) %>%
  group_by(group) %>%
  group_split() %>%
  set_names(., map_depth(., 1, \(x) as.character(first(x$group))))

drc_e2_plates <- imap(e2_references, function(d, nm) {
  assign(nm, d, envir = .GlobalEnv)        
  drm(fold_act ~ dose,
      data = get(nm, envir = .GlobalEnv), 
      fct  = LL.4(names = names_4))
})
```


```{r warning=FALSE, message = FALSE, comment = "", echo = FALSE}
msel_e2_refs <- list()
for (nm in names(drc_e2_plates)) {
  msel_e2_refs[[nm]] <- mselect(drc_e2_plates[[nm]], fctList = candidates)
}
```

```{r}
model_selection_tbl <- imap_dfr(msel_plates, ~ {
  # .x is the matrix returned by mselect(); .y is the plate name
  as_tibble(.x, rownames = "model") %>% 
    mutate(plate = .y, 
           # rename columns to something R‐friendly
           logLik   = `logLik`, 
           AIC      = IC,
           ResVar   = `Res var`,
           LoF_pval = `Lack of fit`) %>%
    select(plate, model, logLik, AIC, ResVar, LoF_pval)
})

best_by_plate <- model_selection_tbl %>%
  group_by(plate) %>%
  filter(!is.na(AIC)) %>%
  slice_min(AIC, n = 1) %>%
  ungroup()


# 3. (optional) pivot longer so you can plot “model vs criterion” across plates
model_selection_long <- model_selection_tbl %>%
  pivot_longer(logLik:LoF_pval, names_to = "criterion", values_to = "value")


model_selection_e2tbl <- imap_dfr(msel_e2_refs, ~ {
  # .x is the matrix returned by mselect(); .y is the plate name
  as_tibble(.x, rownames = "model") %>% 
    mutate(plate = .y, 
           # rename columns to something R‐friendly
           logLik   = `logLik`, 
           AIC      = IC,
           ResVar   = `Res var`,
           LoF_pval = `Lack of fit`) %>%
    select(plate, model, logLik, AIC, ResVar, LoF_pval)
})

best_by_e2_plate <- model_selection_e2tbl %>%
  group_by(plate) %>%
  filter(!is.na(AIC)) %>%
  slice_min(AIC, n = 1) %>%
  ungroup()


# 3. (optional) pivot longer so you can plot “model vs criterion” across plates
model_selection_e2_long <- model_selection_e2tbl %>%
  pivot_longer(logLik:LoF_pval, names_to = "criterion", values_to = "value")
```


```{r}
# e.g. boxplot of AIC by model
model_selection_e2_long %>%
  filter(criterion == "AIC") %>%
  ggplot(aes(model, value)) +
    geom_boxplot() +
    coord_flip() +
    labs(title="Distribution of E2-only AIC across plates", y="AIC", x="Candidate model")


model_selection_long %>%
  filter(criterion == "AIC") %>%
  ggplot(aes(model, value)) +
    geom_boxplot() +
    coord_flip() +
    labs(title="Distribution of AIC across plates", y="AIC", x="Candidate model")
```

## Apply the Best Fit Model

```{r}
data_pooled <- esr_fit %>%
  select(
    group = group_pooled,
    log_dose,
    dose,
    fold_act
  ) %>%
  mutate(n_doses = n_distinct(dose), .by = "group") %>%
  filter(n_doses > 2) %>%
  select(-n_doses) %>%
  group_by(group) %>%
  group_split() %>%
  set_names(., map_depth(., 1, \(x) as.character(first(x$group))))

pooled_drc <- map(data_pooled, \(x) drm(
  formula = fold_act ~ dose, data = x, fct = LL.4(names = names_4)
))

plates_drc <- map(data_plate, \(x) drm(
  formula = fold_act ~ dose, data = x, fct = LL.4(names = names_4)
))

```


```{r}
pooled_coefs <- map(pooled_drc, \(x) coef(x)) %>%
  enframe(name = "group_pooled", value = "coefs") %>%
  unnest_wider(coefs) %>%
  rename_with(~str_replace_all(., ":(.+)", "_treatment"))


plates_coefs <- map(plates_drc, \(x) coef(x)) %>%
  enframe(name = "group_plate", value = "coefs") %>%
  unnest_wider(coefs) %>%
  rename_with(~str_replace_all(., "upper", "Dmax")) %>%
  rename_with(~str_replace_all(., ":(.+)", "_treatment_plate"))
```


### Generate E2 Dmax Values


```{r}
e2_drc <- map(e2_references, \(x) drm(
  formula = fold_act ~ dose, data = x, fct = LL.4(names = names_4)
))
```


```{r}
e2_coefs <- map(e2_drc, \(x) coef(x)) %>%
  enframe(name = "plate", value = "coefs") %>%
  unnest_wider(coefs) %>%
  rename_with(~str_replace_all(., ":(.+)", "_e2"))
```


## Add Other Stats for Plotting

```{r}
esr_meta <- esr_data %>%
  distinct(
    labs_genus,
    labs_er,
    receptor,
    plate,
    group_plate,
    group_pooled,
    groups_treatment,
    labs_treatment,
    treatment,
    dose_type
  )
```

```{r}
predict_plates <- map(plates_drc, \(x) predict(
  x, 
  data.frame(dose = seq(min(x[["dataList"]][["dose"]]), 
                        max(x[["dataList"]][["dose"]]), length.out = 200)),
  interval = "confidence")) %>%
  map2(plates_drc, \(x, y) bind_cols(as_tibble(x),
                                     tibble(dose = c(seq(min(y[["dataList"]][["dose"]]), 
                                                           max(y[["dataList"]][["dose"]]), 
                                                           length.out = 200)))
                                     )) %>%
  map(\(x) mutate(x, log_dose = log10(dose))) %>%
  imap(\(x, idx) mutate(x, group_plate = paste0(idx))) %>%
  map(\(x) left_join(x, esr_meta, by = "group_plate"))

predict_pooled <- map(pooled_drc, \(x) predict(
  x, 
  data.frame(dose = seq(min(x[["dataList"]][["dose"]]), 
                        max(x[["dataList"]][["dose"]]), length.out = 200)),
  interval = "confidence")) %>%
  map2(pooled_drc, \(x, y) bind_cols(as_tibble(x),
                                     tibble(dose = c(seq(min(y[["dataList"]][["dose"]]), 
                                                           max(y[["dataList"]][["dose"]]), 
                                                           length.out = 200)))
                                     )) %>%
  map(\(x) mutate(x, log_dose = log10(dose))) %>%
  imap(\(x, idx) mutate(x, group_pooled = paste0(idx))) %>%
  map(\(x) left_join(x, select(esr_meta, -c(ends_with("plate"))), by = "group_pooled"))
```

```{r}
fit_plates <- map(plates_drc, \(x) modelFit(x)) %>%
  map(\(x) as_tibble(x[2,])) %>%
  list_rbind(names_to = "group_plate") %>%
  rename_with(~paste0("mfit_plate_", .), !group_plate) %>%
  rename_with(~str_remove_all(., "\\s"))

fit_pooled <- map(pooled_drc, \(x) modelFit(x)) %>%
  map(\(x) as_tibble(x[2,])) %>%
  list_rbind(names_to = "group_pooled") %>%
  rename_with(~paste0("mfit_pooled_", .), !group_pooled) %>%
  rename_with(~str_remove_all(., "\\s"))
```


````{r}
esr_fit <- esr_data %>%
  left_join(e2_coefs, by = "plate") %>%
  rowwise() %>%
  mutate(rel_resp = (fold_act/upper_e2)*100) %>%
  ungroup() %>%
  left_join(plates_coefs, by = "group_plate") %>%
  left_join(pooled_coefs, by = "group_pooled") %>%
  left_join(fit_plates  , by = "group_plate") %>%
  left_join(fit_pooled  , by = "group_pooled") %>%
  mutate(across(ends_with("pvalue"), ~round(., 3)))


write_csv(esr_fit, "data/receptors/esr_assays_normalized.csv")
```


```{r}
summary_plates <- map(plates_drc, \(x) summary(x))
```



What you extract from each fit:

- ED₅₀  — the [4th] coefficient of your LL.4 model
- Dₘₐₓ  — the [3rd] coefficient (the upper asymptote)

These two numbers are your primary summary metrics:

- ED₅₀ tells you “which treatment gives half-maximal response at the lowest dose?”
- Dₘₐₓ tells you “which treatment can drive the receptor higher, all the way up to its plateau?”

#### Check Data

```{r}
subplots <- map(data_list, \(x) plot_raw(x$data_resid[[1]], x$treat_label, x$annot_resid))
panel_set <-  
    map_depth(subplot_names, 3, \(x) subplots[x]) %>%
    map_depth(2, \(x) list_flatten(x))
  
  dr_plot_list(df, pooled = FALSE, version = "raw") %>% 
  map_subplots(version = "raw") %>%
 grid_panelset(type = "nav_pill")
```


```{r}
panels <- navset_pill_list(!!!panel_set, widths = c(2, 10))
```




```{r}
bgal <- data_check_box(assays, 
                      "treat_name", 
                      "b_gal", 
                      "treat_name")
bgal
```


### Reference Normalization

**Percent of 1 nM E2:** To compare efficacy across very different treatments, it’s useful to express responses as a percentage of a strong reference agonist. [Tubbs et al.](https://app.paperpile.com/my-library/Tubbs-et-al-2016-pD1_6J11GDuiYTdWdyXeiYw) normalized all data to the response of 1 nM 17β-estradiol (E2) on the same receptor, defining that as 100% activation. We will do the same.

>From [Tubbs et al.](https://app.paperpile.com/my-library/Tubbs-et-al-2016-pD1_6J11GDuiYTdWdyXeiYw):
> *Luciferase activity of treatments relative to vehicle-only treatment and normalized to b-galactosidase activity was used to calculate fold receptor activation. All data were normalized to fold activation of a 1 nM E2 treatment and expressed as a percentage of that response as reported previously.*

```{r}
esr_normalized <- assays %>%
  rowwise() %>%
  mutate(normalized = luciferase/b_gal,
         plate      = as.character(plate)) %>%
  ungroup() %>%
  mutate(mean_veh_plate = mean(normalized[treat_type == "vehicle"]), .by = "plate") %>%
  rowwise() %>%
  mutate(fold_act = normalized / mean_veh_plate) %>%
  ungroup() %>%
  mutate(mean_E2_plate = mean(fold_act[ treat=="e2" & dose==1e-9 ]), .by = "plate") %>%
  rowwise() %>%
  mutate(perc_e2 = (fold_act/mean_E2_plate)*100) %>%
  ungroup()
```

#### Check Data

```{r}
normalized <- data_check_box(esr_normalized, 
                             "treat_name", 
                             "normalized", 
                             "treat_name")
normalized
```

```{r}
fold_act <- data_check_box(esr_normalized, 
                             "treat_name", 
                             "fold_act", 
                             "treat_name")
fold_act
```

```{r}
perc_e2 <- data_check_box(esr_normalized, 
                             "treat_name", 
                             "perc_e2", 
                             "treat_name")
perc_e2
```


Now `max_e2` represents “*percent activation relative to 1 nM E2*” for each measurement. A value of 100 means the treatment produces an effect equal to E2 (at that dose), >100 means a stronger effect than E2, and <100 a weaker effect. For treatments tested at multiple doses, their maximum `max_e2` (at the highest dose or plateau) indicates their efficacy relative to E2 (this will be our `Dmax` if a curve is fit).  
  
We will use `fold_act` for curve fitting (since it’s on a fold-change scale anchored at 1 for baseline), and use `perc_e2` mostly for visualization (so that we can plot responses on a intuitive percentage scale). Both scales are useful: fold-change is better for modeling relative increases, while percent-of-E2 is better for comparing different compounds’ strengths.

### Plate‐wise E2 dose–response Dmax (upper asymptote)

Fit a 4-parameter curve to only our E2 data on each plate, extract its upper asymptote (the `Dmax`), and use that as an alternative to calculate `perc_e2`.

– **Pros**: smooths over an outlier single well; we’re anchoring to the actual plateau of the E₂ curve.
– **Cons**: requires a successful E₂ curve fit on every plate (if a plate’s E₂ data are too noisy, we’ll get NAs).

```{r}
treat_vars <- c(
  "treat_type",
  "treat_subtype",
  "treat_name",
  "treat_desc",
  "treat_form",
  "dose_type",
  "treat"
)

plate_e2_vars <- c(
  "plate",
  "date",
  "genus", 
  "subj", 
  "receptor",
  "model",
  "column",
  treat_vars
)
```


```{r}
plate_e2_curves <- esr_normalized %>%
  filter(treat=="e2") %>%
  nest(.by = plate_e2_vars)  %>%
  mutate(
    drm_result = map(data, \(x) drm(
      fold_act ~ dose,
      data = x,
      fct  = LL.4(names = c("Slope", "Lower Limit", "Upper Limit", "ED50"))
    )),
    drm_summary = map(drm_result, \(x) summary(x, od = FALSE))) %>%
  mutate(Dmax   = map(drm_result, \(x) coef(x)[[3]]),
         slope  = map(drm_result, \(x) coef(x)[[1]]),
         Dmin   = map(drm_result, \(x) coef(x)[[2]]),
         ED50   = map(drm_result, \(x) coef(x)[[4]]))
```


```{r}
esr_norm_dmax <- esr_normalized %>%
  left_join(select(plate_e2_curves, plate, Dmax), by="plate") %>%
  rowwise() %>%
  mutate(perc_e2_Dmax = 100 * (fold_act / Dmax)) %>%
  ungroup()  %>%
  filter(!is.na(dose) & !is.na(fold_act)) %>%
  select(
    plate,
    subj,
    genus,
    receptor,
    model,
    treat_type,
    treat_subtype,
    treat = treat_form,
    treat_name,
    treat_desc,
    log10_dose,
    dose,
    replicate,
    fold_act,
    perc_e2,
    perc_e2_Dmax
  ) %>%
  mutate(n_doses = n_distinct(dose), .by = c(plate, treat)) %>%
  filter(n_doses > 2) %>% select(-n_doses)
```

```{r}
percDmax <- data_check_box(esr_norm_dmax, 
                             "treat_name", 
                             "perc_e2_Dmax", 
                             "treat_name")
save_html(percDmax, "visuals/receptor_assays/rel_activation_dmax.html")
percDmax
```


For most cases, `perc_E2_Dmax` produces a relatively equivalent value, but in some of our plates with extremely high or low values, this method seemed to level out our estimates, so we should proceed using `perc_E2_Dmax` as our relative value for comparing results across plates.

```{r}
write_csv(esr_norm_dmax, "data/receptors/esr_assays_normalized.csv")
```


