---
title: "Data Visualization for Bat Sex Ratio Project"
author: "Thomas Raad"
output:
  html_document:
    theme:
      bootswatch: litera
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: "show"
    fig_caption: true
    df_print: kable
params:
  sampleset: "bats"
                     
---

# Set working directory {.tabset}

## Your Version

```{r}
setwd("C:/Users/raadt/OneDrive - University of Nebraska at Omaha/Datasets/Sex_Ratio")
```

## My Version

- I recommend you use the here package instead for directory control.

```{r}
library(here)

i_am("BatSexRatioDataVisualization.Rmd")
```

- Now R will recognize the directory containing that script as the parent of any other paths you reference.

- I also prefer to load some of my most commonly used packages at the beginning of a script. You can do it directly in R Markdown or run an outside R script that loads them all. I do the latter just so that I can keep a master script that loads those packages and reference it from different scripts.

```{r}
source(here("setup/setup_forThomas.R"))
# note that I include here at the beginning to tell the here package to find this file relative to the source directory I set up with the i_am command in the previous chunk.
```

here is the text of that file for ease of copying and pasting:

```{r}
library(bslib)
library(conflicted)
library(here)
library(kableExtra)
library(knitr)
library(paletteer)
library(plotly)
library(reactable)
library(reactablefmtr)
library(rmarkdown)
library(scales)
library(showtext)
library(skimr)
library(thematic)
library(tidyverse)
library(tippy)
library(usethis)
library(rmdformats)
```



# Load Data {.tabset}

## Your Version

### read file

```{r}
df <- read.csv("NC_netting.csv")
```

#### cleanup

```{r}
class(df$SEX)
df$SEX<-as.factor(df$SEX)
df$REPRO_STAT<-as.factor(df$REPRO_STAT)
df2$REPRO_STAT<-as.factor(df2$REPRO_STAT)

df1$X..GRTS.Cell.Id<-as.numeric(df1$X..GRTS.Cell.Id)

summary(df$SEX)
summary(df$REPRO_STAT)
summary(df2$REPRO_STAT)

df[,c(1,3)]
df1<-df[,c(1,7)]

```


## My Version

I am going to load the file and do your cleanup steps in one chunk to reduce the number of individual objects hanging out in your environment.

```{r}
species_tags <- c(
 coto = "Corynorhinus townsendii virginianus" ,
 cora = "Corynorhinus rafinesquii rafinesquii",
 cora = "Corynorhinus rafinesquii macrotis"   ,
 epfu = "Eptesicus fuscus"                    ,
 lano = "Lasionycteris noctivagans"           ,
 labo = "Lasiurus borealis"                   ,
 laci = "Lasiurus cinereus"                   ,
 lain = "Lasiurus intermedius floridanus"     ,
 lase = "Lasiurus seminolus"                  ,
 myau = "Myotis austroriparius"               ,
 mygr = "Myotis grisescens"                   ,
 myle = "Myotis leibii leibii"                ,
 mylu = "Myotis lucifugus"                    ,
 myse = "Myotis septentrionalis"              ,
 myso = "Myotis sodalis"                      ,
 noid = "noid"                                ,
 nyhu = "Nycticeius humeralis"                ,
 pesu = "Perimyotis subflavus"                ,
 pesu = "pesu"                                ,
 tabr = "Tadarida brasiliensis"               ,
 lano = "Lasionycteris notivagans"            
)
```



```{r warning=FALSE}
nc_net <- read_csv(here("data/bats/NC_netting.csv")) %>%
  dplyr::filter(AGE != "Age") %>%
  rename(cell = "| GRTS Cell Id") %>%
  mutate(START_TIME   = as.character(str_glue("{DATE}", "{START_TIME}" , .sep = " ", .na = "")), 
         STOP_TIME    = as.character(str_glue("{DATE}", "{STOP_TIME}"  , .sep = " ", .na = "")),
         CAP_TIME     = as.character(str_glue("{DATE}", "{CAP_TIME}"   , .sep = " ", .na = "")),
         cell         = as.numeric(cell),
         BAT_SP       = case_when(BAT_SP %in% c("unknown", "Unknown", "species") ~"noid",
                                  is.na(BAT_SP) ~"pesu", .default = BAT_SP)) %>%
  mutate(species = fct_recode(BAT_SP, !!!species_tags),
         across(c(START_TIME, STOP_TIME, CAP_TIME), ~mdy_hms(.)),
         DATE = mdy(DATE),
         MDY  = make_date(year = YEAR, month = MONTH, day = DAY)) %>%
  separate_wider_delim(BAT_SP, " ", names = c("BAT_GEN", "BAT_SPE", "BAT_SUB"), too_few = "align_start") %>%
  mutate(across(where(is.character), ~str_remove_all(., "[^\\w+|\\s]"))) %>%
  mutate(across(c(SEX, REPRO_STAT, BAND, AGE), ~str_to_lower(.))) %>%
  mutate(across(c(SEX, REPRO_STAT, AGE), ~replace_na(., "unknown"))) %>%
  mutate(SEX = case_when(SEX == "m" | SEX == "male" ~"male", 
                          SEX == "f" | SEX == "female" ~"female",
                          SEX == "unknown" ~"unknown"),
         AGE = case_match(AGE, 
                          "adult" ~"adult", 
                          "juvenile" ~"juvenile",
                          "jv"       ~ "juvenile",
                          "unknown" ~"unknown")) %>%
  mutate(AGE = fct(AGE, levels = c("juvenile", "adult", "unknown")),
         REPRO_STAT = str_replace_all(REPRO_STAT, " ", "_")) %>%
  mutate(across(c(SEX, REPRO_STAT), ~as_factor(.))) %>%
  select(
    cell,
    STATE,
    species,
    BAT_GEN,
    BAT_SPE,
    BAT_SUB,
    DATE,
    START_TIME, 
    STOP_TIME, 
    CAP_TIME,
    BAND,
    SEX,
    AGE,
    REPRO_STAT,
    RFA_MM,
    MASS_G,
    WING_SCORE
  ) %>%
  arrange(DATE, STATE, START_TIME)

write.table(nc_net, here("data/bats/NC_netting_clean.tsv"), sep = "\t", row.names = F)
```


```{r warning=FALSE}
nc_skim <- skim(nc_net)
nc_skim
```

```{r}
summary_stats <- nc_net %>%
  arrange(species, SEX, AGE, REPRO_STAT, STATE, DATE) %>%
  summarize(
    count     = n(),
    mean_RFA  = mean(RFA_MM),
    mean_MASS = mean(MASS_G),
    .by = c(species, SEX, AGE, REPRO_STAT, STATE, DATE)
  )

summary_sex_state <- summary_stats %>%
  summarize(
    count = sum(count),
    mean_RFA  = mean(mean_RFA),
    mean_MASS = mean(mean_MASS),
    .by = c(SEX, STATE)
  )
```


### Data meeting revisions

```{r}
df <- df[, c(1, 5, 6, 7, 8, 14, 30)]
```

#df[is.na(df),] <- -1


# turn "DATE" into month/day/year

```{r}
df <- separate(df, col = "DATE", c("month", "day", "year"), sep = "/")
```


# Remove excel descriptor row

```{r}
df <- subset(df, df$AGE != "Age")
```

#to see how many recaptures in df 

```{r}
df$BAND <- as.factor(df$BAND)
summary(df$BAND)
```

# SEX term cleanup

```{r}
summary(df$SEX)
df_male <- subset (df, df$SEX=="male" | df$SEX=="Male" | df$SEX=="M")
df_female <- subset (df, df$SEX=="Female")
df_male$SEX <- "male"
df_female$SEX <- "female"
```



#generates dataframe with missing, unknown, etc.

```{r}
df_new <- rbind(df_male, df_female)

colnames(df_new[1]) <- "cell"
```

# gets rid of awkward GRID cell title, makes new column "cell" within df_new

```{r}
df_new$cell <- df_new$X..GRTS.Cell.Id
df_new <- df_new[, c(2:10)]
df_new$cell <- as.factor(df_new$cell)
```

# makes year column numeric

```{r}
df_new$year <- as.numeric(df_new$year)
```


# creates a column where year is a character

```{r}

df_new$year_char <- as.factor(df_new$year)
```



#summarizes BAT_SP, relabels BAT_SP entries as factors

```{r}
summary(df_new$BAT_SP)
df_new$BAT_SP <- as.factor(df_new$BAT_SP)

```



#creates new column "species", sets all to 0

```{r}
df_new$species <- 0
```



#labels unknowns to "noid"(omitting NA cells), and single na to "pesu" within species column

```{r}
df_new[!is.na(df_new$BAT_SP) & (df_new$BAT_SP == "unknown" | df_new$BAT_SP == "Unknown"), "species"] <- "noid"
df_new[is.na(df_new$BAT_SP),]$species <- "pesu"

```



#adds species tags in "species" column


```{r}
df_new$species[df_new$BAT_SP=="Corynorhinus townsendii virginianus"] <- "coto"
df_new$species[df_new$BAT_SP=="Corynorhinus rafinesquii rafinesquii"] <- "cora"
df_new$species[df_new$BAT_SP=="Corynorhinus rafinesquii macrotis"] <- "cora"
df_new$species[df_new$BAT_SP=="Eptesicus fuscus"] <- "epfu"
df_new$species[df_new$BAT_SP=="Lasionycteris noctivagans"] <- "lano"
df_new$species[df_new$BAT_SP=="Lasiurus borealis"] <- "labo"
df_new$species[df_new$BAT_SP=="Lasiurus cinereus"] <- "laci"
df_new$species[df_new$BAT_SP=="Lasiurus intermedius floridanus"] <- "lain"
df_new$species[df_new$BAT_SP=="Lasiurus seminolus"] <- "lase"
df_new$species[df_new$BAT_SP=="Myotis austroriparius"] <- "myau"
df_new$species[df_new$BAT_SP=="Myotis grisescens"] <- "mygr"
df_new$species[df_new$BAT_SP=="Myotis leibii leibii"] <- "myle"
df_new$species[df_new$BAT_SP=="Myotis lucifugus"] <- "mylu"
df_new$species[df_new$BAT_SP=="Myotis septentrionalis"] <- "myse"
df_new$species[df_new$BAT_SP=="Myotis sodalis"] <- "myso"
df_new$species[df_new$BAT_SP=="Nycticeius humeralis"] <- "nyhu"
df_new$species[df_new$BAT_SP=="Perimyotis subflavus"] <- "pesu"
df_new$species[df_new$BAT_SP=="Tadarida brasiliensis"] <- "tabr"
df_new$species[df_new$BAT_SP=="Lasionycteris notivagans"] <- "lano"
```



#creates df2, a "Female"-only dataframe

```{r}
df2<-subset(df,df$SEX=="Female")

df_new$species <- as.factor(df_new$species)
summary(df_new$species)

```


####Eastern Red#####

```{r}
df_labo <- subset(df_new, species=="labo")
summary(df_new$cell)
```


#creates df for labo using agg function and renaming columns

```{r}
df_labo_agg <- aggregate(df_labo$SEX, by=list(df_labo$SEX, df_labo$month, df_labo$day, df_labo$year_char, df_labo$cell), FUN = length)
names(df_labo_agg)[1]<-"sex"
names(df_labo_agg)[2]<-"month"
names(df_labo_agg)[3]<-"day"
names(df_labo_agg)[4]<-"year"
names(df_labo_agg)[5]<-"cell"
names(df_labo_agg)[6]<-"count"

```


#creates df3, making male/female count columns

```{r}
df3 <- spread(df_labo_agg, sex, count)
df3[is.na(df3)]<- 0

```

#makes new column "total", which totals all bats caught (male+female)

```{r}
df3$total <- df3$male + df3$female
```


#making wide instead of long; looking at totals by year(df4), or cell(df5)

```{r}
df4 <- xtabs(count ~ sex + year, data=df_labo_agg)
df5 <- xtabs(count ~ sex + cell, data=df_labo_agg)
```


#read file

```{r}
df_table <- read.csv("table.csv")
```


#makes df6 from df_table with only needed columns

```{r}
df6 <- df_table[, c(2, 14, 15)]
df7 <- merge(df3, df6, by.x = "cell", by.y = "GRTS_ID")
df7$male_prop <- df7$male / df7$total

```

#convert df7 to csv

```{r}
write.csv(df7, "NC_labo.csv", row.names = FALSE)
```


#fit <- glm(data = dataframe you want to test, dep_var~indep~var, family = data distribution)

```{r}
fit <- glm(data=df7, male_prop~centroid_x, family = "quasibinomial")
fit2 <- glm(data=df7, male_prop~centroid_y, family = "quasibinomial")
summary(fit)
summary(fit2)
df7$year1 <- as.integer(levels(df7$year))[df7$year]
fit <- glm(data=df7, male_prop~year1, family = "quasibinomial")
summary(fit)
```

#X and Y are SIGNIFICANT -- Females more abundant south-east corner of NC

####Big Brown#####

```{r}
df_epfu <- subset(df_new, species=="epfu")
```


#creates df for epfu using agg function and renaming columns

```{r}
df_epfu_agg <- aggregate(df_epfu$SEX, by=list(df_epfu$SEX, df_epfu$month, df_epfu$day, df_epfu$year_char, df_epfu$cell), FUN = length)
names(df_epfu_agg)[1]<-"sex"
names(df_epfu_agg)[2]<-"month"
names(df_epfu_agg)[3]<-"day"
names(df_epfu_agg)[4]<-"year"
names(df_epfu_agg)[5]<-"cell"
names(df_epfu_agg)[6]<-"count"

```



#creates df8, making male/female count columns for epfu

```{r}
df8 <- spread(df_epfu_agg, sex, count)
df8[is.na(df8)]<- 0

```


#makes new column "total", which totals all bats caught (male+female)

```{r}
df8$total <- df8$male + df8$female
```


#making wide instead of long; looking at totals by year(df9), or cell(df10)

```{r}
df9 <- xtabs(count ~ sex + year, data=df_epfu_agg)
df10 <- xtabs(count ~ sex + cell, data=df_epfu_agg)
```

#makes df11 from df_table with only needed columns

```{r}

df11 <- df_table[, c(2, 14, 15)]
df12 <- merge(df8, df11, by.x = "cell", by.y = "GRTS_ID")
df12$male_prop <- df8$male / df8$total
write.csv(df12, "NC_epfu.csv", row.names = FALSE)

```


#fit <- glm(data = dataframe you want to test, dep_var~indep~var, family = data distribution)

```{r}
fit <- glm(data=df12, male_prop~centroid_x, family = "quasibinomial")
fit2 <- glm(data=df12, male_prop~centroid_y, family = "quasibinomial")
summary(fit)
summary(fit2)
df12$year1 <- as.integer(levels(df12$year))[df12$year]
fit <- glm(data=df12, male_prop~year1, family = "quasibinomial")
summary(fit)

```



#Y-dimension is NOT significant for epfu -- more male skewed towards east

####Little Brown#####

```{r}
df_mylu <- subset(df_new, species=="mylu")

```


#creates df for mylu using agg function and renaming columns

```{r}
df_mylu_agg <- aggregate(df_mylu$SEX, by=list(df_mylu$SEX, df_mylu$year_char, df_mylu$cell), FUN = length)
names(df_mylu_agg)[1]<-"sex"
names(df_mylu_agg)[2]<-"year"
names(df_mylu_agg)[3]<-"cell"
names(df_mylu_agg)[4]<-"count"

```


#creates df13, making male/female count columns for mylu

```{r}
df13 <- spread(df_mylu_agg, sex, count)
df13[is.na(df13)]<- 0

```

#makes new column "total", which totals all bats caught (male+female)

```{r}
df13$total <- df13$male + df13$female
```


#making wide instead of long; looking at totals by year(df14), or cell(df15)

```{r}
df14 <- xtabs(count ~ sex + year, data=df_mylu_agg)
df15 <- xtabs(count ~ sex + cell, data=df_mylu_agg)

```


#makes df16 from df_table with only needed columns

```{r}
df16 <- df_table[, c(2, 14, 15)]
df17 <- merge(df13, df16, by.x = "cell", by.y = "GRTS_ID")
df17$male_prop <- df13$male / df13$total

```

#fixing issue with above line

```{r}
nrow(df13)  # Should return 165
nrow(df17)  # Should return 164
df13 <- df13[-161, ]

```



#fit <- glm(data = dataframe you want to test, dep_var~indep~var, family = data distribution)

```{r}
fit <- glm(data=df17, male_prop~centroid_x, family = "quasibinomial")
fit2 <- glm(data=df17, male_prop~centroid_y, family = "quasibinomial")
summary(fit)
summary(fit2)
df17$year1 <- as.integer(levels(df17$year))[df17$year]
fit <- glm(data=df17, male_prop~year1, family = "quasibinomial")
summary(fit)

```

#no significance found

####Northern Long-eared####

```{r}
df_myse <- subset(df_new, species == "myse")
```


# creates df for myse using agg function and renaming columns

```{r}
df_myse_agg <- aggregate(df_myse$SEX, by = list(df_myse$SEX, df_myse$year_char, df_myse$cell), FUN = length)
names(df_myse_agg)[1] <- "sex"
names(df_myse_agg)[2] <- "year"
names(df_myse_agg)[3] <- "cell"
names(df_myse_agg)[4] <- "count"

```


# creates df18, making male/female count columns for myse

```{r}
df18 <- spread(df_myse_agg, sex, count)
df18[is.na(df18)] <- 0

```

# makes new column "total", which totals all bats caught (male + female)

```{r}
df18$total <- df18$male + df18$female

```

# making wide instead of long; looking at totals by year (df19), or cell (df20)

```{r}
df19 <- xtabs(count ~ sex + year, data = df_myse_agg)
df20 <- xtabs(count ~ sex + cell, data = df_myse_agg)

```


# makes df21 from df_table with only needed columns

```{r}
df21 <- df_table[, c(2, 14, 15)]
df22 <- merge(df18, df21, by.x = "cell", by.y = "GRTS_ID")
df22$male_prop <- df18$male / df18$total

```

# fit models for male proportion using GLM with quasibinomial distribution

```{r}
fit <- glm(data = df22, male_prop ~ centroid_x, family = "quasibinomial")
fit2 <- glm(data = df22, male_prop ~ centroid_y, family = "quasibinomial")
summary(fit)
summary(fit2)

```


```{r}
df22$year1 <- as.integer(levels(df22$year))[df22$year]
fit <- glm(data = df22, male_prop ~ year1, family = "quasibinomial")
summary(fit)

```


#no signifiance found

####summary table LABO####

```{r}
df7 %>%
  group_by(centroid_x, centroid_y) %>%
  summarise(mean(male_prop, na.rm=TRUE), sd(male_prop, na.rm=TRUE), n=371)

```



####playing with ggplot####

```{r}
ggplot(data=df7, aes(x=centroid_x, y=male_prop, fill=male_prop))
geom_point(size = 3)
theme_minimal()

```



