---
title: "Demographic Data Summaries for AZA Pygmy Loris Provisional SSP"
author: "Alicia Rich"
output:
  html_document:
    theme:
      bootswatch: litera
      css: "Studbook.css"
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: "show"
    fig_caption: true
    df_print: paged
params:
  sampleset: "loris"
  dataset: "culi"
                     
---

```{r global, message = FALSE}
global             <- config::get(config = "default")

here::i_am("DemographicsLorises.Rmd")
source(here::here(global$setup))

source(here(path$AZAstudbooks$functions))
source(here(path$AZAstudbooks$add))
source(here(path$AZAstudbooks$reactables))
```

# Read in Data

```{r, warning = FALSE}
alive <- read.csv(here(path$AZAstudbooks$living25)) %>% 
  filter(Current.Status == "Alive") %>% 
  pull(Studbook.ID) %>% c(2638)

locations <- bind_rows(read.csv(here(path$AZAstudbooks$institutions$current25)), 
                       read.csv(here(path$AZAstudbooks$institutions$current21)), 
                       read.csv(here(path$AZAstudbooks$institutions$historic21)),
                       loc.extra) %>% clean.locations() %>% enhance.locations()

living.25   <- read.csv(here(path$AZAstudbooks$living25))   %>% 
  clean_studbook(., alive, locations)

living.21   <- read.csv(here(path$AZAstudbooks$living21))   %>% 
  clean_studbook(., alive, locations) %>%
  filter(!(ID %in% pull(living.25, ID)))

historic.21 <- read.csv(here(path$AZAstudbooks$historic21)) %>% 
  clean_studbook(., alive, locations) %>%
  filter(!(ID %in% pull(living.25, ID))) %>%
  filter(!(ID %in% pull(living.21, ID)))

```

# Build Studbook

```{r}
studbook <- bind_rows(living.25, 
                      living.21, 
                      historic.21, 
                      missing) %>%
  mutate(TypeEvent = fct(TypeEvent, 
                         levels = c("Birth", "Transfer", "End")),
         across(c(Sire, Dam), ~ na_if(., 0))) %>%
  arrange(ID, OrderLoc, TypeEvent) %>%
  distinct() %>%
  group_by(ID) %>%
  mutate(StartLoc = if_else(TypeEvent != "End", Date, lag(Date)),
         EndLoc   = if_else(TypeEvent == "End", Date, lead(Date))) %>%
  select(
    ID,
    Status,
    OrderLoc,
    Location,
    TypeEvent,
    StartLoc,
    Date,
    EndLoc,
    NameLoc,
    Country,
    Sex,
    Sire,
    Dam
  ) %>% ungroup()


end.records <- studbook %>%
  select(ID,
         Status,
         OrderLoc,
         Location,
         TypeEvent,
         StartLoc,
         Date,
         EndLoc
  ) %>%
  slice_tail(n = 1, by = ID) %>%
  filter(TypeEvent != "End") %>%
  mutate(Date = if_else(Status == "A", today(), NA),
         TypeEvent = "End") %>% 
  mutate(EndLoc = Date) %>%
  select(-Status) %>% distinct()
```

## Organize by Dates

```{r, warning = FALSE}
timeline <- studbook %>%
  filter(TypeEvent == "Birth" & !is.na(Sire)) %>%
  select(ID      = Sire,
         OffspID = ID,
         Location,
         Date) %>%
  bind_rows(select(filter(studbook, TypeEvent == "Birth" & !is.na(Dam)),
                   ID      = Dam,
                   OffspID = ID,
                   Location,
                   Date)) %>%
  left_join(select(studbook,
                   ID,
                   OrderLoc,
                   Location,
                   StartLoc,
                   EndLoc
                   ), by = join_by(ID, Location)) %>%
  mutate(check = if_else(
    between(Date, StartLoc, EndLoc) | 
      is.na(StartLoc) & Date < EndLoc | 
      is.na(EndLoc) & Date > StartLoc | 
      (is.na(StartLoc) & is.na(EndLoc)) | is.na(Date),
    "keep",
    "discard"
  )) %>% filter(check == "keep") %>%
  mutate(TypeEvent = "Breed") %>%
  select(-c(check, OffspID)) %>% 
  bind_rows(end.records) %>%
  distinct() %>% 
  arrange(ID, 
          OrderLoc,
          Date) %>%
  bind_rows(select(studbook,
                   ID,
                   OrderLoc,
                   Location,
                   TypeEvent,
                   StartLoc,
                   Date,
                   EndLoc)) %>% 
  distinct() %>% 
  mutate(TypeEvent = fct(TypeEvent, levels = c(
    "Birth",
    "Transfer",
    "Breed",
    "End"
  ))) %>%
  arrange(ID, 
          OrderLoc,
          TypeEvent,
          Date) %>%
  select(ID,
         OrderLoc,
         Location,
         StartLoc,
         TypeEvent,
         Date,
         EndLoc) %>%
  group_by(ID) %>%
  mutate(Date = case_when(
    is.na(Date) & TypeEvent != "End"  & TypeEvent != "Breed" ~ StartLoc,
    is.na(Date) & TypeEvent == "End"  ~ EndLoc,
    is.na(Date) & TypeEvent == "Birth" ~ lead(Date) - years(1),
    is.na(Date) & TypeEvent == "Transfer" & lead(TypeEvent) != "End" ~
      lag(Date) + days(floor(as.numeric(as.period((interval(lag(Date), lead(Date))), unit = "days"), "days")/2)),
    .default = Date
  )) %>%
  mutate(Date = if_else(
    is.na(Date) & TypeEvent == "Transfer" & lead(TypeEvent) != "End",
    lag(Date) + days(floor(as.numeric(as.period((interval(lag(Date), lead(Date))), unit = "days"), "days")/2)),
    Date
  )) %>%
  mutate(Date = if_else(
    is.na(Date) & lag(TypeEvent) == "Birth" & is.na(lag(Date)),
    lead(Date) - years(1), Date
  )) %>%
  mutate(Date = case_when(
    is.na(Date) & TypeEvent == "Birth" ~ lead(Date) - years(1),
    is.na(Date) & TypeEvent == "End" & !(ID %in% alive) ~ lag(Date) + years(1),
    .default = Date)) %>%
  mutate(
    StartLoc = case_when(
      TypeEvent == "Breed" ~ lag(StartLoc),
      TypeEvent != "End" & TypeEvent != "Breed" ~ Date,
      TypeEvent == "End" & lag(TypeEvent) != "Breed" ~ lag(Date),
      .default = StartLoc
      ),
    EndLoc = case_when(
      TypeEvent == "Breed" & Location == lead(Location) ~ lead(EndLoc),
      lead(TypeEvent) == "Transfer" | lead(TypeEvent) == "End" ~ lead(Date),
      TypeEvent == "End" ~ Date,
      .default = EndLoc
      )
    ) %>% group_by(ID, Location) %>%
  fill(StartLoc, EndLoc, .direction = "updown") %>%
  ungroup()
```

## Revise Studbook

```{r, warning = FALSE}
studbook.revised <- timeline %>%
  filter(TypeEvent != "Breed") %>%
  left_join(select(
    studbook,
    ID,
    Status,
    OrderLoc,
    Location,
    NameLoc,
    Country,
    Sex,
    Sire,
    Dam
  ), by = join_by(ID, OrderLoc, Location)) %>% distinct() %>%
  mutate(DateBirth = if_else(TypeEvent == "Birth", Date, NA)) %>%
  group_by(ID) %>% fill(DateBirth) %>% ungroup()

```

## Create Nested List by Months

```{r, warning = FALSE}
location.dates <- studbook.revised %>%
  select(
    ID,
    Sex,
    Location,
    DateBirth,
    StartLoc,
    EndLoc
  ) %>% distinct()

births.deaths <- studbook.revised %>%
  select(
    ID,
    Status,
    Sex,
    Date = StartLoc,
    TypeEvent
  ) %>%
  filter(TypeEvent != "Transfer") %>%
  pivot_wider(
    id_cols     = c("ID", "Status", "Sex"),
    names_from  = "TypeEvent",
    values_from = "Date"
  )
```


```{r, warning = FALSE}
monthly.census.location <- pmap(
  list(
    pull(location.dates, StartLoc),
    pull(location.dates, EndLoc),
    pull(location.dates, DateBirth),
    pull(location.dates, Location),
    pull(location.dates, Sex),
    pull(location.dates, ID)
  ), census_by_month) %>%
  bind_rows() %>%
  filter(Location != "UND") %>%
  group_by(Date, Location) %>%
  summarise(Individuals = list(
    tibble(ID, Sex, Age)), 
    .groups = "drop") %>%
  group_by(Date) %>%
  summarise(Locations = split(Individuals, 
                              Location), 
            .groups = "drop") %>%
  split(.$Date) %>%
  map(~ .x$Locations)

monthly.census.total <- pmap(
  list(
    pull(location.dates, StartLoc),
    pull(location.dates, EndLoc),
    pull(location.dates, DateBirth),
    pull(location.dates, Sex),
    pull(location.dates, ID)
  ), census_by_month) %>%
  bind_rows() %>%
  group_by(Date) %>%
  summarise(Individuals = list(
    tibble(ID, Sex, Age)), 
    .groups = "drop") %>%
  group_by(Date) %>%
  summarise(Locations = split(Individuals, 
                              Location), 
            .groups = "drop") %>%
  split(.$Date) %>%
  map(~ .x$Locations)
```


```{r, warning = FALSE}
find.sires <- whoisthedaddy(studbook.revised, monthly.census)
find.dams  <- whoisthemommy(studbook.revised, monthly.census)
```

```{r}
final.studbook <- studbook.revised %>%
  mutate(Sire = case_match(ID, 
                           1047 ~ 1012, 
                           2645 ~ 1131, .default = Sire)) %>%
  add.hypotheticals(c(1045, 1046), "Sire") %>%
  add.hypotheticals(c(2423, 2424), "Sire") %>%
  add.hypotheticals(c(2108), "Sire") %>%
  add.hypotheticals(c(2717), "Dam") %>%
  add.hypotheticals(c(2717), "Sire")

write.table(final.studbook, here(path$AZAstudbooks$working), sep = "\t", row.names = F)
```


```{r}

visual.studbook <- final.studbook %>%
  group_by(ID) %>%
  mutate(Age  = calculate_age(DateBirth, Date),
         Year = year(Date)) %>%
  left_join(select(locations, -Label), by = join_by(
    Location == LocAbbrev,
    Country,
    NameLoc
  )) %>%
  mutate(
        color      = case_match(Sex, 
                            "F" ~ colors$f, 
                            "M" ~ colors$m, 
                            "U" ~ colors$u),
        Status      = case_match(Status,
                                 "D" ~ "Deceased",
                                 "A" ~ "Alive",
                                 "H" ~ "Hypothetical"
                                 ),
         across(c(Sire, Dam), ~ replace_na(., 0)),
         colorEvent = case_match(TypeEvent,
                            "Birth"    ~ "#3B7C70FF",
                            "Transfer" ~ "#CE9642FF",
                            "End"      ~ "#898E9FFF",
         )) %>%
  select(
    Status    ,
    ID        ,
    Sire      ,
    Dam       ,
    Location  ,
    TypeEvent ,
    Date      ,
    NameLoc   ,
    Country   ,
    DateBirth ,
    Sex       ,
    Age       ,
    EndLoc    ,
    StartLoc  ,
    OrderLoc  ,
    colorLoc  ,
    iconLoc   ,
    color     ,
    Year      ,
    colorEvent
  ) %>%
  studbook.react(., studbook.cols(.), groupBy = "Status")

visual.studbook
```

# Plotting

## Time Series Plots of Demographic Trends

We first extract monthly population counts grouped by sex and event type.

```{r}
monthly_byLocation <- bind_rows(
  map_dfr(names(monthly.census), function(date) {
    locations <- monthly.census[[date]]
    
    bind_rows(map_dfr(names(locations), function(location) {
      individuals <- locations[[location]][[1]] 
      tibble(Date           = as.Date(date),
             Location       = location,
             Population     = nrow(individuals),
             Males          = sum(individuals$Sex == "M", na.rm = TRUE),
             Females        = sum(individuals$Sex == "F", na.rm = TRUE),
             Undetermined   = sum(individuals$Sex == "U", na.rm = TRUE))
    }))
  })
) %>%
  filter(between(Date, ymd("1983-2-1"), ymd("2024-3-1"))) %>%
  rename(Month = Date)

monthly_Total <- bind_rows(
  map_dfr(names(monthly.census), function(date) {
    locations <- monthly.census[[date]]
    
    bind_rows(map_dfr(names(locations), function(location) {
      individuals <- locations[[location]][[1]] 
      tibble(Date           = as.Date(date),
             Location       = location,
             Population     = nrow(individuals),
             Males          = sum(individuals$Sex == "M", na.rm = TRUE),
             Females        = sum(individuals$Sex == "F", na.rm = TRUE),
             Undetermined   = sum(individuals$Sex == "U", na.rm = TRUE))
    }))
  })
) %>%
  filter(between(Date, ymd("1983-2-1"), ymd("2024-3-1"))) %>%
  rename(Month = Date)
```


```{r}
monthly_Total <- monthly_byLocation %>%
  group_by(Month) %>%
  summarize(
    Population   = sum(Population  ),
    Males        = sum(Males       ),
    Females      = sum(Females     ),
    Undetermined = sum(Undetermined)
  ) %>% ungroup() %>% 
  mutate(SexRatio = round(Males / (Males + Females), digits = 3))

demographics_byLocation <- studbook.revised %>%
  filter(between(Date, ymd("1983-2-1"), ymd("2024-3-1")) & ID < 10000 & TypeEvent != "Transfer" & Location != "UND") %>%
  mutate(Month = floor_date(Date, "months"),
         TypeEvent = if_else(TypeEvent == "End", "Deaths", "Births")) %>%
  group_by(Month, Location, TypeEvent) %>%
  summarize(Count = n()) %>% ungroup() %>%
  pivot_wider(
    id_cols     = c("Month", "Location"),
    names_from  = "TypeEvent",
    values_from = "Count"
  ) %>%
  arrange(Month) %>%
  mutate(across(where(is.numeric), ~ replace_na(., 0)))

demographics_Total <- studbook.revised %>%
  filter(between(Date, ymd("1983-2-1"), ymd("2024-3-1")) & ID < 10000 & TypeEvent != "Transfer") %>%
  mutate(Month = floor_date(Date, "months"),
         TypeEvent = if_else(TypeEvent == "End", "Deaths", "Births")) %>%
  group_by(Month, TypeEvent) %>%
  summarize(Count = n()) %>% ungroup() %>%
  pivot_wider(
    id_cols     = c("Month", "Location"),
    names_from  = "TypeEvent",
    values_from = "Count"
  ) %>%
  arrange(Month) %>%
  mutate(across(where(is.numeric), ~ replace_na(., 0))) %>%
  left_join(monthly_Total, by = join_by(Month)) %>%
  mutate(FertilityRate = round(Births/Females   , digits = 3),
         MortalityRate = round(Deaths/Population, digits = 3))

demographics_byLocation <- demographics_byLocation %>%
  left_join(monthly_byLocation, by = join_by(Month, Location)) %>%
  mutate(FertilityRate = round(Births/Females   , digits = 3),
         MortalityRate = round(Deaths/Population, digits = 3),
         Year          = year(Month)) %>%
  group_by(Location) %>%
  mutate(LocationMonths = length(unique(Month)))
```


```{r}
Means_byLocation <- demographics_byLocation %>%
  group_by(Year, Location) %>%
  summarize()
```

```{r, out.width="100%"}
fills           <- map_depth(colors, 1, \(x) gsub("FF", "33", x))
plot.population <- plot_ly(monthly_Total,
                           mode        = "lines",
                           type        = "scatter",
                           stackgroup  = "one",
                           x           = ~Date, 
                           y           = ~Undetermined,
                           name        = "Undetermined",
                           legendgroup = "By Sex",
                           mode        = "lines",
                           line        = list(color = colors$u),
                           fillcolor   = fills$u
                           ) %>%
                 add_trace(x           = ~Date, 
                           y           = ~Females,
                           name        = "Females",
                           legendgroup = "By Sex",
                           mode        = "lines",
                           stackgroup  = "one",
                           line        = list(color = colors$f),
                           fillcolor   = fills$f
                           ) %>%
                 add_trace(x           = ~Date, 
                           y           = ~Males,
                           name        = "Males",
                           legendgroup = "By Sex",
                           mode        = "lines",
                           stackgroup  = "one",
                           line        = list(color = colors$m),
                           fillcolor   = fills$m
                           )   %>%
                 add_trace(x           = ~Date, 
                           y           = ~Population,
                           name        = "Total count",
                           legendgroup = "Total",
                           mode        = "lines",
                           stackgroup  = "one",
                           line        = list(color = colors$emph),
                           fillcolor   = fills$emph
                           )  %>%
                 plotly::layout(showlegend = T, 
                                title      = "Population Growth over Time",
                                yaxis      = list(title = "Count"),
                                xaxis      = list(rangeslider = list(visible = T)))

plot.population <- plot.population %>%
  plotly::layout(
         xaxis = list(zerolinecolor = "#fffff",
                      zerolinewidth = 1,
                      gridcolor     = "#fffff",
                      showgrid = F),
         yaxis = list(zerolinecolor = "#fffff",
                      zerolinewidth = 1,
                      showgrid = F),
         plot_bgcolor = "#ffffff", 
         width        = 900)


saveWidget(plot.population, paste0(path$visuals$demographics, "/PopulationGrowth_Overall.html"))
plot.population
```


```{r}
hline <- function(y = 0, color = colors$emph) {
  list(
    type = "line",
    y0   = y,
    y1   = y,
    yref = "paper",
    x0   = 0,
    x1   = 1,
    line = list(color = color, dash="dot")
  )
}

plot.sexratio <- plot_ly(monthly_Total,
                           type        = "scatter",
                           x           = ~Date, 
                           y           = ~SexRatio,
                           name        = "Male:Female Ratio",
                           mode        = "lines+markers",
                           line        = list(
                             color     = colors$u,
                             width     = 1.5
                             ),
                           marker      = list(
                             color     = fills$u,
                             size      = 5,
                             opacity   = 0.7,
                             line      = list(
                               color   = colors$u,
                               width   = 1.0
                               )
                             )) %>%
  layout(shapes       = list(hline(0.5), 
                             list(type      = "rect",
                                  fillcolor = fills$m,
                                  line      = list(color = colors$emph),
                                  opacity   = 0.5,
                                  y0        = 0.5, 
                                  y1        = 1.0, 
                                  x0        = 0, 
                                  x1        = "2025-4-1"
                                  ), 
                             list(type      = "rect",
                                  fillcolor = fills$f,
                                  line      = list(color = colors$emph),
                                  opacity   = 0.5,
                                  y0        = 0.0, 
                                  y1        = 0.5, 
                                  x0        = 0, 
                                  x1        = "2025-4-1"
                                  )
                             )
         
         ) %>%
                  plotly::layout(
                      xaxis         = list(zerolinecolor = "#fffff",
                      zerolinewidth = 1,
                      gridcolor     = "#fffff",
                      showgrid      = F,
                      range         = list("1983-2-1", "2025-4-1")),
         yaxis = list(zerolinecolor = "#fffff",
                      zerolinewidth = 1,
                      showgrid      = F,
                      range         = list(0.25, 0.75)),
         plot_bgcolor = "#ffffff", 
         width        = 900)

saveWidget(plot.sexratio, paste0(path$visuals$demographics, "/PopulationSexRatio.html"))

plot.sexratio
```

### Birth Rates

#### Crude Birth Rate (CBR):

- This is a measure of the number of live births per 1,000 N in a population over a year. 
  - CBR = (Number of live births / Estimated midyear Population) * 1,000 

#### Midyear Population:

- This is the estimated average population for the year, typically taken around July 1st. 
  - Example:
    - If there were 672 live births in a population of 107,619, the crude birth rate would be (672/107,619) * 1000 = 6.25 per 1,000 people 

#### Other Birth Rates:

- **General Fertility Rate:**
  - This is the number of live births per 1,000 females. 
- **Age-Specific Birth Rate:**
  - This is the number of live births to females in a specific age group per 1,000 females in that age group. 
- **Singleton Delivery Rate:**
  - This is the number of singleton live births divided by all live births, multiplied by 100. 

```{r}
fertility_rate <- studbook.revised %>%
  filter(TypeEvent == "Birth" & Date > "1983-2-1") %>%
  count(Date       = floor_date(Date, "month")) %>%
  rename(Births    = n)  %>%
  right_join(filter(monthly_Total, Date < "2022-3-1"), by = join_by(Date)) %>%
  arrange(Date) %>%
  mutate(Births        = replace_na(Births, 0)) %>%
  mutate(FertilityRate = Births/Females, Year = year(Date)) %>%
  group_by(Year) %>%
  summarize(
    Males          = mean(Males),
    Females        = mean(Females),
    Undetermined   = mean(Undetermined),
    SexRatio       = mean(SexRatio) ,
    Births         = mean(Births),
    FertilityRate  = mean(FertilityRate)
  ) %>% ungroup() %>% mutate(across(where(is.numeric), ~round(., digits = 2)))

mortality_rate <- studbook.revised %>%
  filter(TypeEvent == "End" & between(Date, "1983-2-1", "2022-3-1") & ID < 10000) %>%
  count(Date       = floor_date(Date, "month")) %>%
  rename(Deaths    = n)  %>%
  right_join(filter(monthly_Total, between(Date, "1983-2-1", "2022-3-1")), by = join_by(Date)) %>%
  arrange(Date) %>%
  mutate(Births        = replace_na(Births, 0)) %>%
  mutate(FertilityRate = Births/Females, Year = year(Date)) %>%
  group_by(Year) %>%
  summarize(
    Males          = mean(Males),
    Females        = mean(Females),
    Undetermined   = mean(Undetermined),
    SexRatio       = mean(SexRatio) ,
    Births         = mean(Births),
    FertilityRate  = mean(FertilityRate)
  ) %>% ungroup() %>% mutate(across(where(is.numeric), ~round(., digits = 2)))
```


```{r}
birth_rate <- studbook.revised %>%
  filter(TypeEvent == "Birth" & Date > "1983-2-1") %>%
  count(DateBirth  = floor_date(DateBirth, "month")) %>%
  rename(Births    = n)

monthly_birthrate <- left_join(monthly_Total, 
                               birth_rate, 
                               by = c("Date" = "DateBirth")) %>%
  mutate(BirthRate = Births / lag(Population, default = first(Population)))
```


```{r}
ggplot(monthly_birthrate, aes(x = Date, y = BirthRate)) +
  geom_line(color = "#68855CFF", size = 1) +
  labs(title = "Birth Rate Over Time",
       x     = "Year",
       y     = "Births per Individual") +
  theme_minimal()
```

### Mortality Rate by Month

To compute mortality rate, count deaths per month.

```{r}
mortality_rate <- studbook.revised %>%
  filter(TypeEvent == "End" & Date < "2025-2-1") %>%
  count(DateDeath = floor_date(Date, "month")) %>%
  rename(Deaths = n)

monthly_mortality <- left_join(monthly_birthrate, mortality_rate, by = c("Date" = "DateDeath")) %>%
  mutate(MortalityRate = Deaths / lag(Population, default = first(Population)))

ggplot(monthly_mortality, aes(x = Date, y = MortalityRate)) +
  geom_line(color = "#AF6458FF", size = 1) +
  labs(title = "Mortality Rate Over Time",
       x     = "Year",
       y     = "Deaths per Individual") +
  theme_minimal()
```


### Analyzing Reproductive Skew Over Time for Males and Females

To analyze reproductive skew, we need to:

1.	Exclude founders (individuals whose Sire or Dam is NA).
2.	Count offspring per parent for both males (sires) and females (dams) over time.
3.	Visualize the distribution of reproductive success across individuals.
4.	Assess reproductive skew by location (i.e., which locations contribute most to births).

#### Step 1 Count Offspring Per Parent (Ignoring Founders)

```{r}
offspring_data <- studbook.revised %>%
  filter(!is.na(Sire) & !is.na(Dam)) %>%
  select(ID, Sire, Dam, DateBirth, Location)

sire_offspring <- offspring_data %>%
  count(Sire, DateBirth = floor_date(DateBirth, "year")) %>%
  rename(OffspringCount = n)

dam_offspring <- offspring_data %>%
  count(Dam, DateBirth = floor_date(DateBirth, "year")) %>%
  rename(OffspringCount = n)
```

#### Step 2: Visualizing Reproductive Skew for Males (Sires)

```{r}
ggplot(sire_offspring, aes(x = DateBirth, y = OffspringCount, group = Sire)) +
  geom_boxplot(fill = colors$m, alpha = 0.5) +
  labs(title = "Reproductive Skew in Males Over Time",
       x     = "Year",
       y     = "Offspring Count per Male") +
  theme_minimal()
```

#### Step 3: Visualizing Reproductive Skew for Females (Dams)

```{r}
ggplot(dam_offspring, aes(x = DateBirth, y = OffspringCount, group = Dam)) +
  geom_boxplot(fill = colors$f, alpha = 0.5) +
  labs(title = "Reproductive Skew in Females Over Time",
       x = "Year",
       y = "Offspring Count per Female") +
  theme_minimal()
```

####  Step 4: Assessing Reproductive Output by Location

```{r}
location_births <- offspring_data %>%
  count(Location, DateBirth = floor_date(DateBirth, "year")) %>%
  rename(BirthCount = n)

ggplot(location_births, aes(x = DateBirth, y = BirthCount, color = Location)) +
  geom_line(size = 1) +
  labs(title = "Reproductive Output by Location",
       x     = "Year",
       y     = "Number of Births") +
  theme_minimal()
```


