---
title: "Demographic Data Summaries for AZA Pygmy Loris Provisional SSP"
author: "Alicia Rich"
output:
  html_document:
    theme:
      bootswatch: litera
      css: "Studbook.css"
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: "show"
    fig_caption: true
    df_print: paged
params:
  sampleset: "loris"
  dataset: "culi"
                     
---

```{r global, message = FALSE, warning = FALSE}
global             <- config::get(config = "default")

here::i_am("Demography.Rmd")
source(here::here(global$setup))

source(here(path$AZAstudbooks$functions))
source(here(path$AZAstudbooks$btp))

source(here(path$AZAstudbooks$load_data))
source(here(path$AZAstudbooks$reactables))

opts_chunk$set(message = FALSE,
               warning = FALSE,
               echo    = TRUE,
               include = TRUE,
               eval    = TRUE,
               comment = "")

```

# Wrangle Studbook Data

```{r, warning = FALSE}
timeline <- assemble_timeline(studbook) %>%
  mutate(Date = case_when(
    is.na(Date) & TypeEvent == "End"   & ID %in% deaths.21_24 ~ ymd("2023-1-1"), 
    is.na(Date) & TypeEvent == "End"   & ID %in% deaths.24    ~ ymd("2025-1-1"), 
    .default = Date)) %>%  
  fill_dates_timeline()
```

## Revise Studbook

```{r, warning = FALSE}
studbook.revised <- timeline %>%
  filter(TypeEvent != "Breed") %>%
  left_join(select(
    studbook,
    ID,
    Status,
    OrderLoc,
    Location,
    NameLoc,
    Country,
    Sex,
    Sire,
    Dam
  ), by = join_by(ID, OrderLoc, Location)) %>% distinct() %>%
  mutate(DateBirth = if_else(TypeEvent == "Birth", Date, NA)) %>%
  group_by(ID) %>% fill(DateBirth) %>% ungroup()
```

### Create Monthly Census by Location

```{r, warning = FALSE}
monthly.census.location <- census.location(studbook.revised, "month")
```

### Fill Missing Parentage Assignments

```{r, warning = FALSE}
find.sires <- whoisthedaddy(studbook.revised, monthly.census.location)
find.dams  <- whoisthemommy(studbook.revised, monthly.census.location)
```

#### Create Hypothetical IDs and Add Missing Parents

```{r}
final.studbook <- studbook.revised %>%
  mutate(Sire = case_match(ID, 
                           1047 ~ 1012, 
                           2645 ~ 1131, 
                           2738 ~ 2622, .default = Sire)) %>%
  mutate(Dam  = if_else(ID == 2738, 2630, Dam)) %>%
  add.hypotheticals(c(1045, 1046), "Sire") %>%
  add.hypotheticals(c(2423, 2424), "Sire") %>%
  add.hypotheticals(c(2108), "Sire") %>%
  add.hypotheticals(c(2717), "Dam") %>%
  add.hypotheticals(c(2717), "Sire") %>%
  mutate(Cohort = year(DateBirth))

write.table(final.studbook, here(path$AZAstudbooks$working), sep = "\t", row.names = F)

deceased <- filter(final.studbook, Status == "D") %>%
  pull(ID)
```

## Revised Summaries to Carry Forward

```{r}
status.summary <- final.studbook %>%
  mutate(LastDate = if_else(TypeEvent == "End", Date, NA),
         LastLoc  = if_else(TypeEvent == "End", Location, NA),
         FirstLoc = if_else(OrderLoc == 1, Location, NA)) %>%
  group_by(ID) %>%
  fill(LastDate, LastLoc, .direction = "up") %>%
  fill(FirstLoc) %>%
  mutate(maxAge = calculate_age(DateBirth, LastDate)) %>%
  ungroup() %>%
  select(ID,
         Cohort,
         Sex,
         Status,
         maxAge,
         DateBirth,
         Sire,
         Dam,
         FirstLoc,
         LastLoc,
         LastDate
         ) %>%
  distinct()

max_age <- slice_max(status.summary, order_by = maxAge, with_ties = TRUE)

timeline.revised <- assemble_timeline(final.studbook) %>%
  mutate(
    Date  = floor_date(Date, "years"),
    Event = case_when(
    TypeEvent == "End" & ID %in% deceased          ~ "Death", 
    TypeEvent == "Transfer" | TypeEvent == "Breed" ~ TypeEvent,
    .default = NA
    )) %>%
  mutate(Age = year(Date) - Cohort) %>%
  filter(!(is.na(Event) & TypeEvent == "End") & TypeEvent != "Birth") %>%
  select(Cohort, Sex, Event, Date, Age, Location) %>%
  distinct()
```


# Summary Statistics

## Life & Repro Table

```{r}
births.deaths <- timeline.revised %>%
  group_by(Age, Cohort, Sex, Date, Event) %>%
  summarize(Count = n()) %>%
  ungroup() %>%
  pivot_wider(id_cols     = c("Age", 
                              "Cohort", 
                              "Sex", 
                              "Date"),
              names_from  = "Event",
              values_from = "Count",
              values_fill = 0) %>%
  select(
         Cohort,
         Sex,
         Date,
         Age,
         Deaths       = Death, 
         Births       = Breed) %>%
  mutate(Sex = case_match(Sex, 
                          "F" ~ "Female",
                          "M" ~ "Male",
                          "U" ~ "Undetermined")) %>%
  arrange(Cohort, Sex, Age) 

births.deaths.total <- births.deaths %>%
  group_by(Cohort, Date, Age) %>%
  summarize(Deaths = sum(Deaths),
            Births = sum(Births)) %>%
  mutate(Sex = "Total") %>%
  ungroup() %>%
  bind_rows(births.deaths) %>%
  arrange(Cohort, Sex, Age) %>%
  filter(Sex != "Undetermined")
```


```{r}
life.table <- final.studbook %>%
  select(
    ID,
    Sex,
    StartLoc,
    EndLoc,
    TypeEvent
  ) %>% 
  filter(TypeEvent != "Transfer") %>%
  mutate(Date = case_match(TypeEvent, 
                           "Birth" ~ StartLoc, 
                           "End"   ~ EndLoc)) %>%
  select(ID, 
         Sex, 
         TypeEvent, 
         Date) %>%
  pivot_wider(id_cols     = c("ID", "Sex"),
              names_from  = "TypeEvent",
              values_from = "Date") %>%
  distinct() %>%
  census(., "year", "cohort") %>% 
  count_census_grouped(., "cohort") %>%
  mutate(Cohort   = as.integer(Group), .keep = "unused") %>%
  pivot_longer(cols      = c("Males", "Females", "Undetermined", "Total"),
               names_to  = "Sex",
               values_to = "Alive") %>%
  filter(Sex != "Undetermined") %>%
  mutate(Sex = case_match(Sex, 
                          "Females"      ~ "Female",
                          "Males"        ~ "Male",
                          "Total"        ~ "Total"),
         Age = year(Date) - Cohort)  %>%
  cohort_grid(., 21, 1988) %>%
  left_join(births.deaths.total, by = join_by(Cohort, Sex, Age, Date)) %>%
  mutate(across(where(is.numeric), ~replace_na(., 0))) %>%
  arrange(Cohort, Sex, Age)  %>%
  group_by(Cohort, Sex) %>%
  mutate(ax   = if_else(Age == 0, 0, lag(Deaths)),
         N0   = if_else(Age == 0, Alive, NA),
         Risk = (Alive + lead(Alive, default = 0)) / 2) %>%
  fill(N0) %>%
  ungroup() %>%
  mutate(lx = if_else(N0 > 0, Alive/N0, 0)) %>%
  group_by(Cohort, Sex) %>%
  mutate(Lx_life = if_else(Age == max(Age), 0, (lx + lead(lx))/2),
         Px      = if_else(Age == max(Age) | lx == 0, 0, lead(lx)/lx)) %>%
  arrange(Cohort, Sex, desc(Age)) %>%
  mutate(Tx = cumsum(Lx_life)) %>%
  ungroup() %>%
  mutate(ex = if_else(lx > 0, Tx / lx, 0),
         qx = if_else(Risk > 0, Deaths / Risk, 0),
         mx = if_else(Risk > 0, 0.5 * Births / Risk, 0)) %>%
  arrange(Cohort, Sex, Age) %>%
  mutate(Fx    = lx * mx) %>%
  mutate(numT  = Age * Fx) %>%
  group_by(Cohort, Sex) %>%
  mutate(R0   = sum(Fx),
         Tnum = sum(numT)) %>%
  ungroup() %>%
  mutate(T = if_else(R0 > 0, Tnum/R0, 0)) %>%
  mutate(lambda_life = if_else(R0 > 0 & T > 0, R0^(1/T), 0)) %>%
  group_by(Cohort, Sex) %>%
  mutate(mleAll     = mle_all(lx, Age),
         mleAge1    = mle_age1(lx, Age),
         FirstRepro = min(Age[mx>0], na.rm = TRUE),
         LastRepro  = max(Age[mx>0], na.rm = TRUE)) %>%
  arrange(Year) %>%
  mutate(lambda_5 = lambda_5(Alive, Year)) %>%
  ungroup() %>%
  arrange(Cohort, Sex, Age) %>%
  mutate(across(where(is.numeric), ~replace_na(., 0))) %>%
  select(
    Cohort,
    Sex,
    N0,
    Date,
    Year,
    Age,
    Nx = Alive,
    lx,
    Px,
    qx,
    ax,
    Lx_life,
    Risk,
    Tx,
    ex,
    Births,
    mx,
    Fx,
    R0,
    T,
    lambda_life,
    lambda_5,
    mleAll,
    mleAge1,
    FirstRepro,
    LastRepro
  )

write.table(life.table, here(path$AZAstudbooks$lifetable), sep = "\t", row.names = F)
```


# Projections

## Subsetting Cohorts

We are going to compare three different grouped cohorts based on birth years:

1.  2020 - 2024
2.  2005 - 2009
3.  1990 - 1994

And then run three versions of these time-based subsets:

A.  Females Only
B.  Males Only
C.  Pooled Dynamics

## Build a Leslie Matrix

```{r}
test <- filter(life.table, Cohort >= 2015 & Cohort <= 2024 & Sex == "Female")
```


```{r}
A1 <- build_leslie(life.table, 2015, 2024, "Female")
A2 <- build_leslie(life.table, 2005, 2014, "Female")
A3 <- build_leslie(life.table, 1995, 2004, "Female")
B1 <- build_leslie(life.table, 2015, 2024, "Male"  )
B2 <- build_leslie(life.table, 2005, 2014, "Male"  )
B3 <- build_leslie(life.table, 1995, 2004, "Male"  )
C1 <- build_leslie(life.table, 2015, 2024, "Total" )
C2 <- build_leslie(life.table, 2005, 2014, "Total" )
C3 <- build_leslie(life.table, 1995, 2004, "Total" )

```

```{r}
projections <- tibble(
  Sex       = rep(c("Females", "Males", "Pooled"), each       = 3),
  Cohorts   = rep(1:3, length.out = 9),
  lambda    = c(
                lambda(A1), 
                lambda(A2), 
                lambda(A3), 
                lambda(B1), 
                lambda(B2), 
                lambda(B3), 
                lambda(C1), 
                lambda(C2), 
                lambda(C3)
                ),
  genTime  = c(
                generation.time(A1), 
                generation.time(A2), 
                generation.time(A3), 
                generation.time(B1), 
                generation.time(B2), 
                generation.time(B3), 
                generation.time(C1), 
                generation.time(C2), 
                generation.time(C3)
                ),
  netReproRate  = c(
                net.reproductive.rate(A1), 
                net.reproductive.rate(A2), 
                net.reproductive.rate(A3), 
                net.reproductive.rate(B1), 
                net.reproductive.rate(B2), 
                net.reproductive.rate(B3), 
                net.reproductive.rate(C1), 
                net.reproductive.rate(C2), 
                net.reproductive.rate(C3))
) %>%
  arrange(Cohorts, Sex)
```

