---
title: "Loris Microbiome Project Update"
author: "Alicia Rich"
output:
  html_document:
    toc: true
    toc_location: "before"
    toc_depth: 4
    number_sections: false
    toc_float: true
    code_folding: "hide"
    fig_caption: true
editor_options: 
  chunk_output_type: inline
params:
  sampleset: "loris"
                     
---

```{r, include = F}

global      <- config::get(config = "default")
path        <- config::get(config = paste0(params$sampleset))

source(paste0(global$setup))
source(paste0(micro$packages))
source(paste0(micro$functions))
source(paste0(micro$inputs))
source(global$knit_engines)

opts_chunk$set(message = FALSE,
               warning = FALSE,
               echo    = FALSE,
               include = TRUE,
               eval    = TRUE,
               comment = "")

seqruns      <- seqruns %>% keep_at(params$sampleset) %>% list_flatten(name_spec = "")
subject_list <- keep_at(subjects, paste0(params$sampleset)) %>% list_flatten(name_spec = "{inner}")

source(path$metadata$bristol)
source(path$metadata$key)
source(path$metadata$diet)
source(path$metadata$meds)
source(path$metadata$housing)
source(path$metadata$repro)
source(path$metadata$health)

```

```{r}
study_days <- tibble(Date = seq.Date(from = ymd(path$day1), 
                                     to   = ymd(path$last), 
                                     by = "day")) %>%
  mutate(study_day = row_number())

metadata <- read.table(path$metadata$summary, header = T, sep = "\t")  %>%
  select(all_of(names(metadata_key))) %>%
  mutate(across(all_of(date.vars), ~ ymd(.)),
         across(all_of(binary.vars), ~ as.character(.)),
         across(all_of(ids),         ~ str_to_lower(.))) %>%
  mutate(across(all_of(binary.vars), ~ case_when(. == "1" ~ "yes",
                                         . == "0" ~ "no",
                                         .default = .))) %>%
  filter(!is.na(CollectionDate)) %>%
  mutate(across(any_of(replace.unknowns), ~ 
                  if_else(is.na(.), replace_na("unknown"), .))) %>%
  arrange(study_day, Subject)
```

```{r, message = FALSE, warning = FALSE}
collection.schedule <- study_days %>% 
  left_join(select(metadata, SampleID, CollectionDate, Subject), by = join_by(Date == CollectionDate)) %>%
  summarize(N_Samples = n(), .by = c(Date, Subject)) %>%
  ungroup() %>%
  pivot_wider(
    id_cols      = Date,
    names_from   = Subject,
    names_prefix = "samples_",
    values_from  = N_Samples
  ) %>%
  select(-c("samples_NA")) %>%
  mutate(across(-all_of("Date"), ~ as.numeric(str_replace_na(., "0")))) %>%
  rowwise() %>%
  mutate(samples_total = sum(c_across(-all_of("Date")))) %>% 
  ungroup()
```


```{r, message = FALSE, warning = FALSE}
extracts <- collection.schedule %>%
  left_join(select(metadata, ExtractID, CollectionDate, Subject), by = join_by(Date == CollectionDate)) %>%
  summarize(N_Extracts = n(), .by = c(Date, Subject)) %>%
  ungroup() %>%
  pivot_wider(
    id_cols      = Date,
    names_from   = Subject,
    names_prefix = "dnaExtracts_",
    values_from  = N_Extracts
  ) %>%
  select(-c("dnaExtracts_NA")) %>%
  mutate(across(-all_of("Date"), ~ as.numeric(str_replace_na(., "0")))) %>%
  rowwise() %>%
  mutate(dnaExtracts_total = sum(c_across(-all_of("Date")))) %>%
  ungroup() %>%
  left_join(collection.schedule)

sequences <- extracts %>%
  left_join(select(metadata, SequenceID, CollectionDate, Subject), by = join_by(Date == CollectionDate)) %>%
  summarize(N_profiles = n(), .by = c(Date, Subject)) %>%
  ungroup() %>%
  pivot_wider(
    id_cols      = Date,
    names_from   = Subject,
    names_prefix = "mb_profiles_",
    values_from  = N_profiles
  ) %>%
  select(-c("mb_profiles_NA")) %>%
  mutate(across(-all_of("Date"), ~ as.numeric(str_replace_na(., "0")))) %>%
  rowwise() %>%
  mutate(mb_profiles_total = sum(c_across(-all_of("Date")))) %>%
  ungroup() %>%
  left_join(extracts) %>%
  mutate(diet_known             = if_else(Date <= ymd("2024-5-14"), "yes", "no"),
         culi_bristols_recorded = if_else(Date <= ymd("2024-5-26"), "yes", "no"))

# Running a few checks to figure out which data to include

analysis.check <- sequences %>%
  rowwise() %>%
  filter(any(combn(across(ends_with("_total")), 2, FUN = function(x) x[1] != x[2]))) %>%
  ungroup()
# It looks like all samples collected have also been extracted and/or sequenced.


sequence.replicates <- sequences %>%
  filter(mb_profiles_total > samples_total)
# We also have only one sequence per sample, so we can simplify the table for summaries.

data.counts <- sequences %>%
  mutate(month = month(Date, label = TRUE, abbr = TRUE),
         year  = year(Date)) %>%
  mutate(month = str_glue("{month}", ", ", "{year}")) %>%
  select(
    month,
    Date,
    diet_known,
    culi_bristols_recorded,
    starts_with("samples_")
  )
```


# Diet 

## Trial Phases

We have the Nutrition Department's records from sampling Day 1 (October 26, 2023) through February 19, 2024 for the 2-week intervals of different combinations of probiotic (yes/no), oatgel (yes/no), and steroid (low/high dose). After that, I kept notes from keepers when these three items were adjusted on and off through trial and error, but it would be good to verify those records. The table below summarizes our records to date for those diet trials.

```{r}
culi.diet.schedule %>%
  list_rbind() %>%
  mutate(Probiotic   = case_when(diet$probiotic == 1   ~ "check", 
                                 diet$probiotic == 0   ~ "minus",
                                 is.na(diet$probiotic) ~ "question"),
         Oatgel      = case_when(diet$oatgel == 1   ~ "check", 
                                 diet$oatgel == 0   ~ "minus",
                                 is.na(diet$oatgel) ~ "question"),
         Steroid     = case_when(diet$steroid == 0.00 ~ "minus",
                                 diet$steroid > 0.00  ~ as.character(diet$steroid),
                                 is.na(diet$steroid)  ~ "question")) %>%
  select(start_day,
         Steroid,
         Probiotic,
         Oatgel) %>%
  gt() %>%
    fmt_date(columns = ends_with("_day"), date_style = "m_day_year") %>%
    cols_label(start_day = "Begin") %>%
    fmt_icon( columns = Steroid, rows = Steroid == "minus" | Steroid == "question") %>%
    fmt_icon( columns = c("Probiotic", "Oatgel")) %>%
    cols_align("center") %>%
    tab_header("Summary of Trial Phases for Culi") %>%
    tab_footnote("Phase ends on the date the next row begins.", 
                 locations = cells_column_labels(start_day)) %>%
    tab_footnote("Steroid (Budesonide) dose reported as mg/day", 
                 locations = cells_column_labels(Steroid)) %>%
    tab_footnote("Warble stayed at baseline for duration of the study.", 
                 locations = cells_title()) %>%
    opt_stylize(style = 3, color = "gray")

```

## Other Food Items

I tried to note the switch dates when keepers mentioned experimentally removing specific items from Culi's food rotation in an attempt to mitigate GI distress. It would be great to have any more details on these and other food items that are readily available so that we can try out some modeling to look for any responses to those diet changes.

```{r}
tagList(
  tags$h2("Foods Removed for Culi:"),
imap(foods.removed$culi, ~ {
  food  <- .y
  date  <- .x
  header  <- paste0(month(date, label = TRUE, abbr = TRUE), " ", mday(date), ", ", year(date))
  descr   <- paste0(str_to_title(food), " removed from Culi's rotation.")
  tags$p(
    tags$strong(header),
    tags$br(),
      descr
  )
}))
```

# Health Records

## Bristol Fecal Scores

Keepers provided us with their daily records of Culi's bristol fecal scores, so we have included that as an exploratory predictor variable in our analysis (again, using daily means where we have multiple data points per day). Currently our records stop after May 26, 2024, so it would be helpful to include updated records for our other sampling days as well. 

## Ad Lib Notes

I kept ad lib records of any relevant symptomatic information I witnessed or keepers reported to me. It's not really enough to build any inferential statistics from now, but if vet staff would like to include any readily available records like this, we could certainly add it to some of our models to try link symptoms with gut biomarkers.

```{r}
tagList(
  tags$h2("Ad Lib Symptom Notes for Culi:"),
imap(health.records$culi, ~ {
  date  <- .y
  note  <- .x
  header  <- paste0(month(date, label = TRUE, abbr = TRUE), " ", mday(date), ", ", year(date))
  tags$p(
    tags$strong(header),
    tags$br(),
      tags$em(note)
  )
}))
```

## Medications

Keepers mentioned administering antibiotics to Culi once during our sample collection, so I noted the medication name, dosage, and dates. If he received antibiotics or antiparasitics any other time during our sample collection, we should definitely be sure to include that in any reports/publications, and we will need to verify whether this had any significant effect on any microbial patterns in the data. Below is what I have right now.

```{r}
meds$culi %>%
  mutate(med_type = str_to_title(med_type),
         total_days = last_day - start_day) %>%
  select(med_type,
         start_day,
         med_name,
         med_dose,
         last_day,
         total_days) %>%
  filter(med_type != "None") %>%
  gt(rowname_col = "med_type") %>%
  tab_header(title = "Medications administered to Culi") %>%
  tab_footnote("Steroid for diet trials not included", locations = cells_title()) %>%
  tab_footnote("Medication dose reported as mg/day", locations = cells_column_labels(med_dose)) %>%
  cols_label(start_day  = "Day Started",
             last_day   = "Day Completed",
             med_dose   = "Dose Administered",
             med_name   = "Medication Name",
             total_days = "Total Days") %>%
  fmt_date(columns = c("start_day", "last_day"), date_style = "yMd") %>%
  opt_stylize(style = 3, color = "pink")
  
```




# Sampling Snapshot

Our first day of sample collection was on October 26, 2023. The table below summarizes how many fecal daily samples we have for each individual beginning on that date. Each time a sample is collected, we have to run it through a series of pipeline phases in the lab before we produce a complete microbiome profile generated from our onsite MinION sequencers. To date, we have complete profiles for every sample included in this table. In cases where we generated quality data from more than one sample per day for a subject, we essentially merged those samples to generate mean values so that our dataset's finest temporal scale looks at daily averages for each individual.

```{r}
reactable(
    data.counts,
    groupBy = "month",
    defaultPageSize = 15,
    height = 600,
    resizable = TRUE,
    columns = list(
    month                   = colDef(name   = "Months", footer = "N Total:"),
    Date                    = colDef(name = "Days", sticky = "left", aggregate = "count",
                                     footer = JS("function(column, state) {
        let total = 0
        state.sortedData.forEach(function(row) {
          total += row[column.id]
        })
        return total + ' Days';
      }")
    ),
    diet_known              = colDef(name = "Diet Trial",
                                     cell = function(value) {if (value == "no") "\u274c No" else "\u2714\ufe0f Yes"},
                                     aggregate = "frequency",
  aggregated = JS("
  function(cellInfo) {
        const values = cellInfo.subRows.map(function(row) { 
        return row['diet_known'] === 'no' ? '\u274c No' : '\u2714\ufe0f Yes' 
        })
      // Count occurrences of each value
      const counts = values.reduce(function(acc, v) {
        acc[v] = (acc[v] || 0) + 1;
        return acc;
      }, {});
      
      // Format the counts as a string
      return Object.entries(counts)
        .map(([key, count]) => `${key}: ${count}`)
        .join(', ');
  }
  ")
  ),
    culi_bristols_recorded  = colDef(name = "Culi's Bristol Scores", 
                                     cell = function(value) {
    if (value == "no") "\u274c No" else "\u2714\ufe0f Yes"
  },
                                     aggregate = "frequency",
  aggregated = JS("
  function(cellInfo) {
        const values = cellInfo.subRows.map(function(row) { 
        return row['culi_bristols_recorded'] === 'no' ? '\u274c No' : '\u2714\ufe0f Yes' 
        })
      
      // Count occurrences of each value
      const counts = values.reduce(function(acc, v) {
        acc[v] = (acc[v] || 0) + 1;
        return acc;
      }, {});
      
      // Format the counts as a string
      return Object.entries(counts)
        .map(([key, count]) => `${key}: ${count}`)
        .join(', ');
  }
  ")
  ),
    samples_culi            = colDef(name = "Culi"           , aggregate = "sum",
                                     footer = JS("function(column, state) {
        let total = 0
        state.sortedData.forEach(function(row) {
          total += row[column.id]
        })
        return total + ' Samples for Culi';
      }")
      ),
    samples_warble          = colDef(name = "Warble"         , aggregate = "sum",
                                     footer = JS("function(column, state) {
        let total = 0
        state.sortedData.forEach(function(row) {
          total += row[column.id]
        })
        return total + ' Samples for Warble';
      }")),
    samples_unknown         = colDef(name = "Unknown"        , aggregate = "sum",
                                     footer = JS("function(column, state) {
        let total = 0
        state.sortedData.forEach(function(row) {
          total += row[column.id]
        })
        return total + ' Samples without Subject ID';
      }")),
    samples_total           = colDef(name = "Total"          , aggregate = "sum",
                                     footer = JS("function(column, state) {
        let total = 0
        state.sortedData.forEach(function(row) {
          total += row[column.id]
        })
        return total + ' Total Samples Analyzed';
      }")
      )
  ),
  columnGroups = list(
    colGroup(name = "Sampling Schedule", columns = c("month", "Date")),
    colGroup(name = "Records Available", columns = c("diet_known", "culi_bristols_recorded")),
    colGroup(name = "N Samples Analyzed"  , columns = c("samples_culi", "samples_warble", "samples_unknown", "samples_total")
    )
  ),
  defaultColDef = colDef(footerStyle = list(fontWeight = "bold"))
)
```

