---
title: "Receptor Assay Data Tidying"
author: "Alicia Rich"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: 
      bslib: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(forcats)
library(conflicted)
library(plotly)
library(htmltools)
library(htmlwidgets)
library(drc)
library(shiny)
library(bslib)
library(flexdashboard)
library(rlang)
library(crosstalk)
library(paletteer)
library(broom)
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = ""
  )

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(plotly::layout)


source("data/receptors/inputs.R")
source("helpers/dose_response_curves/variables_helpers.R")
source("helpers/dose_response_curves/model_options.R")
source("helpers/dose_response_curves/models_results.R")
source("helpers/dose_response_curves/tidy_data.R")
source("helpers/dose_response_curves/plot_traces.R")
source("helpers/dose_response_curves/widget_plots.R")
source("helpers/dose_response_curves/widget_layouts.R")
source("helpers/dose_response_curves/map_widgets.R")
source("helpers/dose_response_curves/widget_dashboards.R")

esr_data <- read_csv("data/receptors/esr_assays_normalized.csv")

```

```{r eval=FALSE, include=FALSE}
#working color palette

c("#BDA14DFF", 
  "#3EBCB6FF", 
  "#0169C4FF", 
  "#153460FF", 
  "#D5114EFF", 
  "#A56EB6FF", 
  "#4B1C57FF", 
  "#09090CFF")

c("#BDA14D80", 
  "#3EBCB680", 
  "#0169C480", 
  "#15346080", 
  "#D5114E80", 
  "#A56EB680", 
  "#4B1C5780", 
  "#09090C80")

c("#BDA14D33", 
  "#3EBCB633", 
  "#0169C433", 
  "#15346033", 
  "#D5114E33", 
  "#A56EB633", 
  "#4B1C5733", 
  "#09090C33")
```

# Create factors for modeling

```{r}
model_data <- esr_data %>%
  mutate(across(c(
    subj,
    receptor,
    plate,
    treat_subtype,
    treat
  ), ~factor(.))) %>%
  select(subj,
         receptor,
         model,
         plate,
         treat_subtype,
         treat,
         log10_dose,
         dose,
         fold_act
         ) %>%
  group_by(model) %>%
  group_split() %>%
  set_names(., map(., \(x) unique(pull(x, "model"))))
```

```{r}
data_by_treat <- esr_data %>%
  mutate(group = as.character(str_glue("{model}_{treat}"))) %>%
  mutate(across(c(
    group,
    subj,
    receptor,
    plate,
    model,
    treat_subtype,
    treat
  ), ~factor(.))) %>%
  select(subj,
         receptor,
         model,
         treat_subtype,
         treat,
         group,
         log10_dose,
         dose,
         fold_act
         ) %>%
  group_by(group) %>%
  group_split() %>%
  set_names(., map(., \(x) as.character(unique(pull(x, "group")))))
```

```{r}
models_by_treat <- map_depth(data_by_treat, 1, \(x) drm_base(x))
```

```{r}
select_by_treat <- map_depth(models_by_treat, 1, \(x) drm_select(x))
```


```{r}
select_byplate <- map(
  model_data, \(x) mselect(x, fctList = list(
    LL.2(),
    LL.3(),    
    LL.4(),    
    W1.4(),    
    W2.4()
    ))
  )
```


```{r}
result <- model_safely(model_data,  drm_default,  ~drm_default(.x, TRUE))

safe_summary <- safely(summary)

result_summary <- map(result, \(x) safe_summary(x)) %>%
  map_depth(1, \(x) compact(keep_at(x, "result"))) %>%
    compact() %>%
    list_flatten(name_spec = "{outer}")
```





```{r}
dose_response_byplate <- drm(fold_act ~ dose, 
                     curveid = plate_treat,
                     data    = model_data, 
                     fct     = LL.4(names = c("Slope", 
                                              "Lower Limit", 
                                              "Upper Limit", 
                                              "ED50")),
                     separate = TRUE)
```

```{r}
dr_coefs <-  model_coef_tbl(dose_response, "LL4") %>%
  bind_rows(model_coef_tbl(dose_response_byplate, "LL4"))
```

```{r}
compared <- compParm(dose_response, "ED50")
```


```{r}
dr_pooled <- nest(esr_data, .by = group_pooled) %>%
  model_dose_response("linear") %>%
  rename(lm_result  = drm_result, 
         lm_summary = drm_summary) %>%
  model_dose_response("default")
```



# Fitting Dose–Response Models (LL.4 Logistic)

For each treatment that has more than one dose tested, we will fit a four-parameter log-logistic model (LL.4) to the fold-change data. The four parameters of this model are: lower asymptote (minimal effect, often near 1 if baseline-normalized), upper asymptote (maximal effect = Dmax), the inflection point or EC50 (the dose producing half-maximal response), and the slope (Hill coefficient). We use the drc package’s `drm()` function with the `LL.4()` model specification for this task.\n
\n
Grouping by Treatment: We will fit separate curves for each unique combination of receptor, species, and treatment.\n
\n
The model `drm(fold_act ~ dose, fct=LL.4())` will attempt to estimate the four parameters for each curve. *Note: We ensure dose is on a numeric scale (e.g., molar concentrations or mg/mL as appropriate). If dose values vary over many orders of magnitude, using a log scale for fitting might be helpful; however, drc’s LL.4 expects the raw dose and internally handles the log-logistic form, so we provide the raw dose (non-logged) in the formula.*\n
\n

>**After fitting, examine each model’s results:**
>- Use `summary()` on the `drm_result` to see estimated parameters and standard errors.
>- The summary of the curve fitting shows the estimates of each of the four parameters and their standard errors. The p-values tell us if the parameters are different from zero.
>- Check if any models failed to converge or gave unreasonable estimates (e.g., extremely high EC50 or slope). We might detect these by looking at the tidy coefficients or trying to predict outside the data range.

## Step 1: Plate-by-plate QC plots


```{r}
plate_fits <- drm_nested(esr_data, c(
  "plate",
  "genus", 
  "subj", 
  "receptor",
  "model",
  "treat_type",
  "treat_subtype",
  "treat",
  "treat_name",
  "treat_desc"
))
```

## Step 2: Pooled "average" curves

Now we will pool data across plates and just group by the cell model and the treatment for ease of comparison.

```{r}
pooled_fits <- drm_nested(esr_data, c(
  "genus", 
  "subj", 
  "receptor",
  "model",
  "treat_type",
  "treat_subtype",
  "treat",
  "treat_name",
  "treat_desc"
))
```

## Step 3: Compare Visuals


```{r}
plates <- dr_vis_summary(plate_fits, FALSE)

pooled <- dr_vis_summary(pooled_fits, TRUE)

summary <- navset_bar(
  nav_panel("By Plate", plates),
  nav_panel("Plates Pooled", pooled)
)
save_html(summary, "visuals/receptor_assays/model_summary.html")
```


```{r}
save_html(plates, "visuals/receptor_assays/model_summary_byplate.html")
save_html(pooled, "visuals/receptor_assays/model_summary_pooled.html")
```


```{r}
summary
```



